---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r getdata}
zip_file_name = "./activity.zip"

# Unzip data file
unzip(zip_file_name, exdir="./", overwrite=TRUE)
data <- read.csv("./activity.csv")
```


## What is mean total number of steps taken per day?
```{r meansteps}
# Load in dependencies
library(dplyr)
library(ggplot2)

# Group up number of steps by day and calculate total number of steps per day
data_totals <-
  data %>%
  group_by(date) %>%
  summarise(total_steps=sum(steps, na.rm=TRUE))

# Calculate mean and median of total steps per day
mean_steps <- mean(data_totals$total_steps, na.rm=TRUE)
median_steps <- median(data_totals$total_steps, na.rm=TRUE)

# Make histogram of step totals
with(data_totals,
     hist(data_totals$total_steps, breaks=15,
          xlab="Total Steps", main="Histogram of Total Steps Per Day"))

# Plot vertical lines at mean and median
abline(v=mean_steps, col="red", lty=2, lwd=2)
abline(v=median_steps, col="blue", lty=2, lwd=2)

# Add a legend for the Veritcal lines
legend("topright", legend=c("Mean", "Median"),
       col=c("red", "blue"), lty=2, lwd=2)
```

First I grouped the step numbers by the date using group_by and take the sum of the total steps on each day. I then use this to calculate the mean and median steps of each day. For this dataset the mean of the total steps for each day was `r floor(mean_steps)` and the median was `r median_steps`. When plotting the histogram I used 15 break points.

## What is the average daily activity pattern?
```{r dailyactivitypattern}
# Calculate average number of steps at each interval point across all days
activity_avgs <-
  data %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm=TRUE))

# Find interval point that has max number of average steps
max_steps <- max(activity_avgs$mean_steps)
max_interval <- activity_avgs$interval[activity_avgs$mean_steps == max_steps]
msg <- paste("Max:", as.character(floor(max_steps)), "steps")

# Plot average steps 
with(activity_avgs,
     plot(interval, mean_steps, type="n",
          xlab="Interval", ylab="Average Steps"))

title(main="Average Daily Activity")
lines(activity_avgs$interval, activity_avgs$mean_steps)
abline(v=max_interval, col="blue", lty=2, lwd=2)
text(max_interval+330, max_steps-5, msg, col="blue")
```

For this question I started by grouping by the interval for all dates in the dataset and took the mean of the steps at each interval across all dates. I then found the maximum average steps to be `r floor(max_steps)` at interval `r max_interval`. I plotted the time series and added a vertical line at the interval that had the highest average steps.

## Imputing missing values
```{r imputemissingvalues}
library(impute)
library(reshape2)

# Get number of missing values in dataset
total_nas <- sum(is.na(data$steps))

# Get average for every interval in dataset
interval_avgs <- 
  data %>%
  group_by(interval) %>%
  summarize(mean_steps = mean(steps, na.rm=TRUE))

# Make temporary variable for imputed data
imputed_data <- data

# Replace intervals that have missing steps data with average
# Step number at that interval
for (i in 1:dim(data)[1]){
  if(is.na(imputed_data$steps[i])){
    # Get interval of missing row
    interval <- imputed_data$interval[i]
    # Get index in averages data frame
    index <- interval_avgs$interval==interval
    # Replace NA with average at that interval
    imputed_data$steps[i] = interval_avgs$mean_steps[index]
  }
}

# Plot histogram of dataset before and after imputing
par(mfrow=c(1,2))
hist(data$steps, main="Before Imputing", xlab="Steps", col="red")
hist(imputed_data$steps, main="After Imputing", xlab="Steps", col="blue")
```

This dataset had a total of `r total_nas` missing values. I chose to impute this missing data by replacing them with the average step number at the respective interval across all dates. To accomplish this, I grouped the dataset by its interval and took the mean of the steps at each interval. I then used a for loop to look through the steps column for missing values. When a missing value is found, it is replaced with the average value at the corresponding interval. I then plotted a histogram of the dataset before and after imputing.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdayweedkenddiff}
library(lubridate)

# Function that determines if date is a weekday or weekend
weekend_or_weekday <- function(xday){
  weekdays <- c(2,3,4,5)
  lub <- wday(ymd(xday))
  ifelse(lub %in% weekdays, "weekday", "weekend")
}

weekday_means <- 
  data %>%
  # Create new factor column that shows if date is a weekend or weekday
  mutate(weekday = weekend_or_weekday(date)) %>%
  mutate(weekday = as.factor(weekday)) %>%
  # Take mean steps of each interval on weekends and weekdays
  group_by(weekday, interval) %>%
  summarise(mean_steps = mean(steps, na.rm=TRUE))

# Cast to wide format of dataset to make plotting easier
weekday_activity <- dcast(weekday_means, interval~weekday, value.var="mean_steps")

# Plot data
with(weekday_activity, plot(interval, weekday, type="n",
                            ylab="Average steps", xlab="Interval"))

title(main="Average Daily Activity on Weekdays vs Weekends")

lines(weekday_activity$interval, weekday_activity$weekday, col="red")
lines(weekday_activity$interval, weekday_activity$weekend, col="blue")

legend("topright", legend=c("Mean weekday steps", "Mean weekend steps"),
       col=c("red", "blue"), lty=2, lwd=2)
```

To find the weekend/weekday differences I started by using a lubridate function to determine which dates in the dataset were weekdays or weekends. I then group by weekdays and intervals and calculate the means of those groups and plotted the mean number of steps at each interval for weekdays and weekends